---
title: "TMDB 5000 — Exploratory Data Analysis"
author: "Gerard Kenneally"
date: 5/5/26
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: true
    theme: flatly
execute:
  warning: false
  message: false
---

# Overview

This notebook explores the TMDB 5000 Movie Dataset to assess its
suitability for our group project dashboard, document data quality
issues we'll need to handle, and lock in the questions our final
dashboard will answer.

**Data source:** [TMDB 5000 Movie Dataset on Kaggle](https://www.kaggle.com/datasets/tmdb/tmdb-movie-metadata)

# Setup

```{r}
library(tidyverse)
library(jsonlite)
library(scales)

# Custom theme — same one used in the Shiny dashboard
theme_tmdb <- function(base_size = 12) {
  theme_minimal(base_size = base_size) +
    theme(
      text             = element_text(family = "Helvetica", color = "#221F1F"),
      plot.title       = element_text(face = "bold", size = base_size + 4),
      plot.subtitle    = element_text(color = "#555555", size = base_size),
      plot.caption     = element_text(color = "#888888", size = base_size - 2,
                                      hjust = 0),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(color = "#EEEEEE"),
      plot.background  = element_rect(fill = "white", color = NA),
      axis.title       = element_text(color = "#555555"),
      legend.position  = "bottom"
    )
}

theme_set(theme_tmdb())

# Shared color palette
palette_primary   <- "#E50914"  # signal red
palette_secondary <- "#221F1F"  # near-black
palette_neutral   <- "#888888"  # grey for non-data ink

movies_raw  <- read_csv("data/tmdb_5000_movies.csv",  show_col_types = FALSE)
credits_raw <- read_csv("data/tmdb_5000_credits.csv", show_col_types = FALSE)
```

# Shape

How big is the data, and what variables do we have?

```{r}
dim(movies_raw)
dim(credits_raw)
```

```{r}
glimpse(movies_raw)
```

```{r}
glimpse(credits_raw)
```

The movies table has `r nrow(movies_raw)` rows and `r ncol(movies_raw)`
columns. The credits table has `r nrow(credits_raw)` rows and
`r ncol(credits_raw)` columns. They join on `id` ↔ `movie_id`.

This comfortably exceeds the project's minimum of 500 observations
and 5 variables.

# JSON-encoded columns

Several columns store data as JSON-encoded strings rather than
plain text. We'll need to parse these to use them.

```{r}
movies_raw |>
  select(genres, keywords, production_companies, production_countries,
         spoken_languages) |>
  slice(1) |>
  as.list()
```

This is going to require a parsing helper. We've handled this in
`global.R` for the dashboard; for the EDA we'll do a quick parse here.

```{r}
parse_json_names <- function(json_string) {
  if (is.na(json_string) || json_string == "" || json_string == "[]") {
    return(character(0))
  }
  parsed <- tryCatch(fromJSON(json_string), error = function(e) NULL)
  if (is.null(parsed) || length(parsed) == 0) return(character(0))
  parsed$name
}

movies <- movies_raw |>
  mutate(
    release_date  = as.Date(release_date),
    release_year  = as.integer(format(release_date, "%Y")),
    decade        = (release_year %/% 10) * 10,
    genre_names   = map(genres, parse_json_names),
    primary_genre = map_chr(genre_names, ~ if (length(.x) > 0) .x[1] else NA_character_),
    budget_clean  = if_else(budget  == 0, NA_real_, budget),
    revenue_clean = if_else(revenue == 0, NA_real_, revenue),
    profit        = revenue_clean - budget_clean,
    roi           = revenue_clean / budget_clean
  )
```

# Data quality

## Missing values

```{r}
movies |>
  summarize(across(c(budget, revenue, runtime, release_date,
                     vote_average, vote_count),
                   ~ sum(is.na(.x) | .x == 0))) |>
  pivot_longer(everything(), names_to = "variable", values_to = "missing_or_zero") |>
  mutate(pct = round(100 * missing_or_zero / nrow(movies), 1)) |>
  arrange(desc(missing_or_zero))
```

The big quality issue here is that **TMDB stores unknown budget and
revenue as `0`, not `NA`**. A lot of films in this dataset have either
`budget == 0` or `revenue == 0`, which means "we don't know," not
"this movie cost nothing." Any analysis involving budget, revenue,
profit, or ROI needs to filter these out — otherwise we'll get
nonsense.

## Outliers and weirdness

```{r}
summary(movies$budget_clean)
summary(movies$revenue_clean)
summary(movies$runtime)
```

```{r}
# Films with implausible runtimes
movies |>
  filter(runtime < 60 | runtime > 240) |>
  select(title, runtime, release_year) |>
  arrange(runtime)
```

# Distributions

## Release year

```{r}
movies |>
  filter(!is.na(release_year)) |>
  ggplot(aes(x = release_year)) +
  geom_histogram(binwidth = 5, fill = palette_primary, color = "white", linewidth = 0.3) +
  scale_x_continuous(breaks = seq(1920, 2020, 20)) +
  scale_y_continuous(labels = label_comma()) +
  labs(
    title    = "Most films in TMDB are from the last 20 years",
    subtitle = "Distribution of films by release year, 5-year bins",
    x        = NULL,
    y        = "Number of films",
    caption  = "Source: TMDB 5000 Movies dataset"
  )
```

## Budget

```{r}
movies |>
  filter(!is.na(budget_clean)) |>
  ggplot(aes(x = budget_clean)) +
  geom_histogram(bins = 40, fill = palette_primary, color = "white", linewidth = 0.3) +
  scale_x_log10(
    labels = label_dollar(scale_cut = cut_short_scale()),
    breaks = c(1e3, 1e5, 1e7, 1e9)
  ) +
  scale_y_continuous(labels = label_comma()) +
  labs(
    title    = "Budget is roughly log-normal once we drop the unknowns",
    subtitle = "Films with reported budget = 0 are excluded as missing data",
    x        = "Budget (log scale)",
    y        = "Number of films",
    caption  = "Source: TMDB 5000 Movies dataset"
  )
```

## Genre frequency

```{r}
movies |>
  unnest(genre_names) |>
  count(genre_names, sort = TRUE) |>
  ggplot(aes(x = n, y = fct_reorder(genre_names, n))) +
  geom_col(fill = palette_primary) +
  geom_text(aes(label = label_comma()(n)), hjust = -0.2, size = 3, color = palette_secondary) +
  scale_x_continuous(labels = label_comma(), expand = expansion(mult = c(0, 0.1))) +
  labs(
    title    = "Drama and comedy dominate the dataset",
    subtitle = "Films can have multiple genre tags; counts therefore exceed 4,803",
    x        = NULL,
    y        = NULL,
    caption  = "Source: TMDB 5000 Movies dataset"
  ) +
  theme(panel.grid.major.y = element_blank())
```

## Ratings

```{r}
movies |>
  filter(vote_count >= 50) |>
  ggplot(aes(x = vote_average)) +
  geom_histogram(binwidth = 0.2, fill = palette_primary, color = "white", linewidth = 0.3) +
  geom_vline(aes(xintercept = median(vote_average, na.rm = TRUE)),
             color = palette_secondary, linetype = "dashed", linewidth = 0.5) +
  annotate("text", x = 6.4, y = 350,
           label = "Median", hjust = 1.1, size = 3, color = palette_secondary) +
  scale_y_continuous(labels = label_comma()) +
  labs(
    title    = "Most films cluster around a 6.5 rating",
    subtitle = "Films with fewer than 50 votes excluded to reduce noise",
    x        = "Average user rating",
    y        = "Number of films",
    caption  = "Source: TMDB 5000 Movies dataset"
  )
```

# Questions the dashboard will answer

Based on this exploration, our dashboard will focus on the following
questions. Each is supported by the variables present in the dataset
and is interesting enough to motivate interactive filtering.

1.  **How well does budget predict revenue, and how does the
    relationship vary by genre?** Budget-vs-revenue scatter, colored or
    faceted by genre, with year filter.

2.  **Which genres deliver the highest typical return on investment?**
    Median ROI by genre, with filters for decade and minimum vote count
    to control for sample-size effects.

3.  **Has movie length changed over time, and does the trend differ
    between high-budget and low-budget films?** Runtime by release year,
    split by budget tertile.

4.  **Do critic-favored films and audience-popular films overlap, or
    diverge?** Scatter of `vote_average` against `popularity` (or
    `vote_count`), with hover-to-identify outliers.

5.  **Which directors and actors consistently appear in profitable
    films?** Searchable leaderboard of people-level stats, computed
    from the parsed `cast` and `crew` columns.

We'll likely cut one of these by the final dashboard if it doesn't
hold up — five is the upper end of what one dashboard can answer well.

# Ethical considerations

This dataset is about commercial films and the public-figure cast and
crew who made them, so the privacy concerns are minimal. Two issues
worth being upfront about in the final presentation:

- **Coverage bias.** TMDB is a community-maintained database and skews
  toward English-language and Western films. Any "best of" or "most
  successful" framing in the dashboard should be qualified — it's the
  best of what TMDB volunteers chose to catalog, not the global film
  industry.

- **Nominal dollars.** Budgets and revenues are not adjusted for
  inflation. A $10M budget in 1990 is not the same as $10M in 2015.
  We'll either inflation-adjust or be explicit about this limitation
  in the final dashboard.